---
title: "Tidy Tuesday: Statistical Performance Indicators"
subtitle: "Week of 2025-11-25"
author: "Emily C Rutkowski"
format: 
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: true
    code-summary: "Show code"
---

# Introduction

This week's Tidy Tuesday dataset comes from the **World Bank Statistical Performance Indicators (SPI)**. The SPI measures how well countries collect, process, and use statistical data. A higher score means a country has better statistical infrastructure and practices.

**Research Question:** How has the statistical performance of a country changed over time?

## What I Learned: Tabset Panels in Quarto! 🎉

::: {.callout-tip}
## New skill I learned this week!

I learned how to use **tabset panels** in Quarto. Tabset panels organize content into clickable tabs, so readers can switch between different views without scrolling through a long document.

To create a tabset, you use this syntax:

```
::: {.panel-tabset}

## Tab 1 Name
Content for tab 1

## Tab 2 Name
Content for tab 2

:::
```

The `:::` markers create a special "div" container, and `.panel-tabset` tells Quarto to make it into tabs. Honestly, I thinkt this is one of the coolest things about Quarto...
:::

# Load Libraries
```{r}
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(tidytuesdayR)
```

# Load Data 
```{r}
#| message: false
#| warning: false

 
tuesdata<- tidytuesdayR::tt_load('2025-11-25')
# This is the main dataset with overall spi scores by country and year
spi_data <- tuesdata$spi
# view data 

```

## Check years
```{r}
#| message: false
#| warning: false
# see what years are available in the data
# unique() gives all the different values in a column
# sort() puts them in order
sort(unique(spi_data$year))
```

## Check Countries 
```{r}
#| message: false
#| warning: false

# How many countries are in the dataset?
n_distinct(spi_data$country) # n_distinct() counts unique values
```

That a lot of countries. I only want to keep 5. 

```{r}

selected_countries <- c("United States", "Brazil", "India", "Nigeria", "Germany") # create a vector of countries to analyze


spi_subset <- spi_data %>% # filter the data to only include selected countries
  filter(country %in% selected_countries) %>% # checks if country is in my list 
  filter(year >= 2016) %>%
  select(country, year, overall_score) # pick columns to keep 

# Check that it worked
head(spi_subset)
```

## Visualizations {.tabset}

Now for the fun part - let's visualize how statistical performance has changed over time!

::: {.panel-tabset}

## Line Plot 1 
```{r}
# Create a line plot showing SPI over time for each country

line_plot_all <- spi_subset %>%
  ggplot(aes(x = year,             # year 
             y = overall_score,    # SPI score 
             color = country,      # different color for each country
             group = country)) +   # group points by country for the lines
  geom_line(linewidth = 1.2) + # set line thickness
  geom_point(size = 3) + # set points on line for individual data and adjust size of dot 
  scale_x_continuous(breaks = seq(2004, 2023, by = 2)) + # sets which years to show as tick marks
  scale_color_brewer(palette = "Set2") + # "Set2" is a colorblind friendly palette 
  labs(title = "How Has Statistical Performance Changed Over Time?", # add titles and axis labels 
       subtitle = "Comparing 5 countries from different regions (2004-2023)",
       caption = "Data: World Bank SPI | TidyTuesday 2025-11-25",
       x = "Year",
       y = "Statistical Performance Index (SPI)",
       color = "Country") + # set color to individual country 
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"), # edit title
        legend.position = "bottom") # put legend on bottom of plot 

line_plot_all

ggsave(here("Tidy_Tuesday", "Week_04", "output", "spi_over_time.png"),
            width = 10,
            height = 6)
```
 
## Faceted Plot 2
```{r}
# Create a faceted plot - each country gets its own mini-plot
facet_plot <- spi_subset %>%
  ggplot(aes(x = year, # set x and y 
             y = overall_score,
             color = country)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ country, ncol = 2) + # arrange in 2 columns 
  scale_x_continuous(breaks = seq(2016, 2023, by = 2)) + # which years to show on x axis 
  scale_color_brewer(palette = "Set2") +
  labs(title = "Statistical Performance Trends by Country",
       subtitle = "Each country shown in its own panel for detailed view",
       caption = "Data: World Bank SPI | TidyTuesday 2025-11-25",
       x = "Year",
       y = "SPI Score") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"),
        legend.position = "none", # hide legend since colors match the facet labels
        strip.text = element_text(size = 11, face = "bold")) # mke labels stand out 

facet_plot

ggsave(here("Tidy_Tuesday", "Week_04", "output", "spi_over_time_facet.png"),
            width = 10,
            height = 6)
```

:::

# Summary

All five countries show generally improving statistical performance from 2016-2023. Germany and the United States consistently score highest, while Nigeria shows the most room for growth.